Memory for television episodes preserves event content while introducing new across-event similarities

### 2019-12-11 paper/main.tex
started updating text

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors or outdoors; the names of any characters on the screen; the names of any characters who were in focus in the camera shot; the names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 27 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (36 events detected).  \textbf{C.} Average topic vectors for each of the 36 video events. \textbf{D.} Topic vectors for each of 252 sentences spoken by an example participant while recalling the video, resampled to match the timseries of the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (29 events detected).  \textbf{F.} Average topic vectors for each of the 29 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).  The heat maps in each panel were created using \texttt{Seaborn}~\citep{WaskEtal16}.}
\label{fig:model}
\end{figure}

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls, we treated each (overlapping) 10-sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g., Participants P2, P6, and P14), while others' recalls segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appear to be integrating across mutliple events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event~\citep[Figs.~\ref{fig:model}D, \corrmats; also see][]{MannEtal11, HowaEtal12}.

@@

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  Overall, we found that participants heavily clustered video events in their recalls by both temporal proximity (mean: 0.860, SEM: 0.016) and semantic content (mean: 0.888, SEM: 0.014).

% Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.860, SEM: 0.016) and that participants with higher temporal clustering scores also tended to recall more events (Pearson's $r(15) = 0.50, p = 0.042$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.888, SEM: 0.014) and participants who exhibited stronger semantic clustering remembered more video events overall (Pearson's $r(15) = 0.55, p = 0.02$).

>>>>

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors or outdoors; the names of any characters on the screen; the names of any characters who were in focus in the camera shot; the names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (30 events detected).  \textbf{C.} Average topic vectors for each of the 30 video events. \textbf{D.} Topic vectors for each of 265 sliding windows of sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (22 events detected).  \textbf{F.} Average topic vectors for each of the 22 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).  The heat maps in each panel were created using \texttt{Seaborn}~\citep{WaskEtal16}.}
\label{fig:model}
\end{figure}

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls, we treated each (overlapping) 10-sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-windows by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), while others' recalls segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appear to be integrating across mutliple events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event~\citep[Figs.~\ref{fig:model}D, \corrmats; also see][]{MannEtal11, HowaEtal12}.

@@

We can also apply two list-learning-native analyses that describe how participants group items in their recall sequences: temporal clustering and semantic clustering \citep[][see \textit{Methods} for details]{PolyEtal09}.  Temporal clustering refers to the extent to which participants group their recall responses according to encoding position.  Overall, we found that sequentially viewed video events were clustered heavily in participants' recall event sequences (mean: 0.767, SEM: 0.029), and that participants with higher temporal clustering scores also tended to perform better according to both \cite{ChenEtal17}'s hand-annotated memory scores (Pearson's $r(15) = 0.62, p = 0.008$) and our model's estimate (Pearson's $r(15) = 0.54, p = 0.024$).  Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.  We found that participants tended to recall semantically similar video events together (mean: 0.787, SEM: 0.018), and participants who exhibited stronger semantic clustering remembered more video events overall (Pearson's $r(15) = 0.55, p = 0.02$).

% Overall, we found that participants heavily clustered video events in their recalls by both temporal proximity (mean: 0.860, SEM: 0.016) and semantic content (mean: 0.888, SEM: 0.014).

% Semantic clustering measures the extent to which participants cluster their recall responses according to semantic similarity.
